---
title: "Malignant neoplasms of digestive organs C15-C26 in Chile from 2010 to 2023"
author:
  - name: "Valentina Meilan Ip Francesconi"
    email: "valentina.ip@ug.uchile.cl "
    orcid: "0000-0000-0000-0000"
  - name: "Amaru Simón Agüero Jiménez"
    email: "aaguero@miaundes.cl"
    orcid: "0000-0001-7336-1833"
date: "08/11/2024"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    number-depth: 3
    code-fold: true
    bibliography: ref.bib
    csl: apa-numeric-superscript.csl
---

```{css, echo=FALSE}
caption {
  caption-side: top;
  text-align: center;
}
```

# Introduction

Malignant neoplasms of the digestive organs (MNDO) are one of the main causes of health burden and mortality worldwide, representing a growing challenge for healthcare systems in both developed and developing countries [@Wang2023Global]. According to global estimates, in 2019, digestive cancers made a significant contribution to the incidence and mortality of non-communicable diseases, with age-standardized mortality rates reaching up to 102 per 100,000 inhabitants and more than 8 million annual deaths without any significant decrease in global incidence observed over the past decades [@Wang2023Global]. Within this group, colorectal, gastric, liver, and pancreatic cancers stand out due to their high frequency and lethality and are responsible for a substantial proportion of disability-adjusted life years (DALYs) worldwide [@Wang2023Global].

In the Latin American context, specifically in Chile, the burden of digestive neoplasms is particularly significant. National studies have shown that cancers of the digestive organs are the most prevalent in the southern part of the country, and that mortality attributable to cancer increased from 21% in 1990 to 26% in 2017, reflecting the epidemiological and demographic transition the region is experiencing [@LaraMadrid2024Caracterizacion]. Widely documented risk factors include advanced age, male sex, a diet high in salt and processed meats, alcohol and tobacco consumption, obesity, diabetes, Helicobacter pylori infection, and chronic liver diseases such as viral hepatitis [@Wang2023Global]. Moreover, multimorbidity is common in patients with digestive cancer, complicating clinical management, and increasing hospital costs and the probability of complications [@Bernal2022Multimorbidity;@Jiang2025Analyzing].

The distribution of the disease burden is not homogeneous. There are significant regional and sociodemographic differences in the incidence and mortality of digestive cancers associated with environmental factors, access to health services, prevalence of risk factors, and social determinants [@Gonzalez2025Cholecystectomy]. For example, in Chile, mortality from gallbladder cancer is among the highest in the world, especially among women and in southern regions, which has prompted public health interventions such as expanding access to cholecystectomy [@Gonzalez2025Cholecystectomy]. Surgery remains the main therapeutic strategy for resectable digestive tumors; however, it is associated with a significant decrease in postoperative functional capacity and quality of life, which has driven research into pre- and postoperative interventions to improve functional outcomes and patient rehabilitation [@LaraMadrid2024Caracterizacion;@Navarrete2022Calidad]. However, late detection and the lack of effective screening programs in most countries in the region mean that most cases are diagnosed at advanced stages, limiting therapeutic options and reducing overall survival [@Wang2023Global].

In this context, it is essential to provide an updated characterization of the hospital burden attributable to malignant neoplasms of the digestive organs by analyzing its temporal, geographic, age group, and sex distribution. This information is vital for guiding public policy, targeting preventive strategies, and optimizing resource allocation within the healthcare system. The present study aimed to analyze the incidence and trends of hospitalization due to malignant digestive neoplasms in Chile during 2010–2023, with the goal of providing evidence to inform public health decision-making and the planning of more effective and equitable interventions.

## Aim

Assess the incidence of hospitalizations due to malignant neoplasms digestive organs in Chile between the years 2010 and 2023.

# Methodology

The study design used was a historical retrospective cohort based on the analysis of national anonymized secondary data from hospital records between 2010 and 2023. Open data on hospital discharges are available from the Department of Health Statistics and Information of the Ministry of Health of Chile [@DEIS2024]. 
This corresponds to a quantitative design, historical retrospective cohort based on the analysis of secondary, national, anonymized data from hospital records between the years 2010 and 2023. These open data on hospital discharges are available in the Department of Health Statistics and Information of the Ministry of Health of Chile, accessible via the following [link](https://deis.minsal.cl/#datosabiertos) @[DEIS2024]. For the diagnosis of MNOD, codes C15-C26 from the International Classification of Diseases revision 10 corresponding to the chapter on MNOD were used. These begin with malignant neoplasms of the esophagus (C15), followed by those of the stomach (C16). Next were those of the small intestine (C17), colon (C18), rectosigmoid junction (C19), and rectum (C20). Neoplasms of the anus and anal canal (C21) are also considered, as are those of the liver and intrahepatic bile ducts (C22), such as hepatocarcinomas and intrahepatic cholangiocarcinomas. Subsequently, those of the gallbladder (C23) and other parts of the biliary system, such as extrahepatic cholangiocarcinomas (C24), were included. Malignant neoplasms of the pancreas (C25) and, finally, those of other digestive organs or unspecified locations (C26) are also included. These hospital rates will be evaluated according to sociodemographic variables and their temporal evolution between the years 2010 and 2023.

## R packages

```{r}
#| message: false
#| warning: false

install_and_load <- function(package) {
  if (!require(package, character.only = TRUE)) {
    utils::install.packages(package)
    library(package, character.only = TRUE)
  }
}

# List of packages to be installed and loaded
packages <- c("devtools", "renv", "tidyverse", "janitor", "data.table", "flexsurv","readxl",
              "kableExtra", "reticulate", "FactoMineR", "factoextra", "knitr","plotly","censo2017", "ggbreak", "patchwork","latex2exp","chilemapas","biscale", "sf","DT","qrcode")

# Apply the function to each package
invisible(capture.output(sapply(packages, install_and_load)))

opts_chunk$set(
  warning = FALSE,
  message = FALSE
)
rm(list = ls())
```

## Data Administration

```{r, HOSP}

# Define the path for the HOSP.rds file
output_file <- paste0(gsub("docs", "", getwd()), "data/data_output/HOSP.rds")

# Check if the file already exists
if (!file.exists(output_file)) {
  # If the file doesn't exist, execute the code
  # Get the list of CSV files in the "data/EGRESOS" folder
  files <- list.files(path = paste0(gsub("docs", "", getwd()), "data/EGRESOS/"), pattern = "\\.csv$", full.names = TRUE)

  # Define the columns you need to select, considering potential variations in the column names
  required_columns <- c("SEXO", "GRUPO_EDAD", "ETNIA", 
                        "GLOSA_PAIS_ORIGEN", "COMUNA_RESIDENCIA", "GLOSA_COMUNA_RESIDENCIA", 
                        "REGION_RESIDENCIA", "GLOSA_REGION_RESIDENCIA", "PREVISION", "GLOSA_PREVISION", 
                        "ANO_EGRESO", "DIAG1", "DIAG2", "DIAS_ESTADA", "CONDICION_EGRESO")

  # Function to read each file and select the required columns
  read_file <- function(file) {
    # Read the data, assuming the delimiter is a semicolon (you can change it if necessary)
    data <- read_delim(file, delim = ";")
    
    # Select the necessary columns (handling cases where the column name is slightly different)
    data_selected <- data %>%
      select(any_of(required_columns)) 
    
    return(data_selected)
  }

  # Load all datasets and name them according to the file name (without extension)
  data_list <- lapply(files, read_file)

  # Assign names to the list elements using the file names without the extension
  names(data_list) <- tools::file_path_sans_ext(basename(files))

  # Combine all datasets into one
  HOSP <- bind_rows(data_list) %>%
    filter(SEXO != "*")

  # Save the combined dataset as an RDS file
  saveRDS(HOSP, file = output_file)

} else {
   # If the file exists, load it
  HOSP <- readRDS(output_file)
}

HOSP <- HOSP %>%
  mutate(across(where(is.character), ~ iconv(., from = "latin1", to = "UTF-8"))) %>%  # Convert all character columns to UTF-8
  mutate(
    GRUPO_EDAD = as.character(GRUPO_EDAD),  # Ensure GRUPO_EDAD is character
    GRUPO_EDAD = recode(GRUPO_EDAD, 
                        `menor de un año` = "0 - 9", 
                        `1 a 9` = "0 - 9"),  # Recode the age groups
    GRUPO_EDAD = gsub(" a ", " - ", GRUPO_EDAD),
    GRUPO_EDAD = gsub(" y más", " +", GRUPO_EDAD),# Replace "a" with "-"
    GRUPO_EDAD = as.factor(GRUPO_EDAD)  # Convert back to factor if needed
  ) %>% filter(SEXO == "HOMBRE" | SEXO == "MUJER") %>%
  mutate(SEXO = case_when(
    SEXO == "HOMBRE" ~ "Male",
    SEXO == "MUJER" ~ "Female",
    TRUE ~ NA_character_  # Por si hay valores inesperados
  ))  # Filter out unwanted values in SEXO column

```

```{r}
P_DATA <- read_excel(paste0(gsub("docs", "", getwd()), "/data/TOTAL/P_Data_Extract_From_World_Development_Indicators.xlsx"), 
    col_types = c("skip", "skip", "text", 
        "text", "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric")) %>%
  filter(grepl("FE.5Y$", `Series Code`) | grepl("POP.TOTL.FE.IN$", `Series Code`))  %>%
  pivot_longer(cols = starts_with("20"),
               names_to = "Year",  
               values_to = "Population") %>% 
  complete(`Series Code`, Year) %>%
  select(-1) %>% 
  pivot_wider(names_from = `Series Name`, 
              values_from = Population) %>%
  mutate(across(c(-Year,-`Population, female`), ~ . * `Population, female`/100)) %>%
  pivot_longer(cols = -Year,
               names_to = "GRUPO_EDAD",
               values_to = "n") %>%
  mutate(Grupo_Edad_Agrupado = case_when(
    grepl("00-04|05-09", GRUPO_EDAD) ~ "0 - 9",
    grepl("10-14|15-19", GRUPO_EDAD) ~ "10 - 19",
    grepl("20-24|25-29", GRUPO_EDAD) ~ "20 - 29",
    grepl("30-34|35-39", GRUPO_EDAD) ~ "30 - 39",
    grepl("40-44|45-49", GRUPO_EDAD) ~ "40 - 49",
    grepl("50-54|55-59", GRUPO_EDAD) ~ "50 - 59",
    grepl("60-64|65-69", GRUPO_EDAD) ~ "60 - 69",
    grepl("70-74|75-79", GRUPO_EDAD) ~ "70 - 79",
    grepl("80", GRUPO_EDAD) ~ "80 +",
    TRUE ~ "Total"  # This is included as a safeguard in case there are unexpected values
  )) %>%
  group_by(Year, Grupo_Edad_Agrupado) %>%  # Group by Year and the new age group
  summarise(n = sum(n, na.rm = TRUE)) %>% 
  mutate(Year = gsub(" \\[.*\\]", "", Year))  # Remove unwanted characters from the Year column
```

```{r}
# Define the path for the CENSO2017.rds file
output_file_censo <- paste0(gsub("docs", "", getwd()), "data/data_output/CENSO2017.rds")

# Check if the file already exists
if (!file.exists(output_file_censo)) {
  # If the file doesn't exist, download, process, and save the data

  # Connect and download census data
  censo_conectar()
  censo_descargar()
  
  # Select relevant columns and mutate data
  CENSO2017 <- censo_tabla("personas")[,c(1,5,6)]
  CENSO2017 <- CENSO2017 %>%
    mutate(p09 = replace(p09, p09 %in% c(132, 131), NA)) %>%
    mutate(p08 = replace(p08, p08 %in% c(3, 0), NA)) %>%
    mutate(SEXO = factor(p08, levels = c(1, 2), labels = c("Male", "Female"))) %>%
    mutate(GRUPO_EDAD = cut(p09,
                           breaks = c(-Inf, 9, 19, 29, 39, 49, 59, 69, 79, 89, Inf),
                           labels = c("0 - 9", "10 - 19", "20 - 29", "30 - 39", "40 - 49", 
                                      "50 - 59", "60 - 69", "70 - 79", "80 - 89", "90 +"),
                           right = FALSE))
  
  # Save the processed dataset as an RDS file
  saveRDS(CENSO2017, file = output_file_censo)
} else {
  # If the file exists, load it
  CENSO2017 <- readRDS(output_file_censo)
}

STDPOPULATION <- CENSO2017 %>% 
  group_by(GRUPO_EDAD, SEXO) %>%
  summarise(n = n(), .groups = 'drop')  %>%
  mutate(POP_STD = n*100000/sum(n))

```

## Hospitalization Rates of Malignant neoplasms of female genital organs C51-C58 ICD chapter standardized by age group.

### Direct standardization rates (DSR).

For the purpose of calculating the hospitalization rate for each C51-C58 ICD code, the direct standardization by age group (stratum) method was employed [@chiang1984life] [@https://doi.org/10.1002/sim.4780100317]. The total population of Chile by year (2010-2023) is obtained from World Bank Open Data. [@worldbank_population_data]. This method utilizes a standard population from the 2017 Chilean Census as a reference [@censo2017]. The formula used to calculate DSR is as follows:

```{=tex}
\begin{align*}

DSR=\frac{\sum^k_{i=1}N_i\frac{y_i}{n_i}}{\sum^k_{i=1}N_i}

\end{align*}
```
where

$N_i$ is the reference population size for the *i*th stratum.

$y_i$ is the number of events observed in population in the *i*th stratum.

$n_i$ is the person-time for the *i*th stratum of the study population.

```{r}
################################################################################
# ICD-10 Mapping para neoplasias malignas de órganos digestivos (C15–C26)
################################################################################
icd10_mapping <- c(
  "C15" = "Esophagus",
  "C16" = "Stomach",
  "C17" = "Small intestine",
  "C18" = "Colon",
  "C19" = "Rectosigmoid junction",
  "C20" = "Rectum",
  "C21" = "Anus and anal canal",
  "C22" = "Liver and intrahepatic bile ducts",
  "C23" = "Gallbladder",
  "C24" = "Other parts of biliary tract",
  "C25" = "Pancreas",
  "C26" = "Other and ill-defined digestive organs"
)

################################################################################
# YEAR
################################################################################
HOSP_C15_C26 <- bind_rows(
  # Casos anuales por edad y sexo
  HOSP_C15_C26_YEAR <- HOSP %>%
    filter(grepl("^C1[5-9]|^C2[0-6]", DIAG1)) %>%
    group_by(ANO_EGRESO, GRUPO_EDAD, SEXO) %>%
    summarise(n = n(), .groups = 'drop') %>%
    mutate(across(everything(), ~na_if(., NA) %>% replace_na(0))) %>%
    mutate(ANO_EGRESO = as.character(ANO_EGRESO)),
  
  # Total (todos los años) por edad y sexo
  HOSP_C15_C26_TOTAL <- HOSP %>%
    filter(grepl("^C1[5-9]|^C2[0-6]", DIAG1)) %>%
    group_by(GRUPO_EDAD, SEXO) %>%
    summarise(n = n(), .groups = 'drop') %>%
    mutate(across(everything(), ~na_if(., NA) %>% replace_na(0))) %>%
    mutate(ANO_EGRESO = "Total")
)

# Totales por edad, sexo y año (para denominador)
TOTALS_HOSP_C15_C26 <- bind_rows(
  # Total pop (todos los años) por edad y sexo
  TOTALS_HOSP_C15_C26_TOTAL_POP <- HOSP %>%
    group_by(GRUPO_EDAD, SEXO) %>%
    summarise(total_pop = n(), .groups = 'drop') %>%
    mutate(ANO_EGRESO = "Total"),
  
  # Total pop por año, edad y sexo
  TOTALS_HOSP_C15_C26_YEAR_POP <- HOSP %>%
    group_by(GRUPO_EDAD, SEXO, ANO_EGRESO) %>%
    summarise(total_pop = n(), .groups = 'drop')
)

# Cálculo de tasas y DSR por edad y sexo
HOSP_C15_C26_RATES <- HOSP_C15_C26 %>%
  left_join(TOTALS_HOSP_C15_C26, 
            by = c("GRUPO_EDAD", "SEXO", "ANO_EGRESO")) %>%
  # Se asume STDPOPULATION tiene las columnas: GRUPO_EDAD, SEXO, POP_STD
  left_join(STDPOPULATION[, c("GRUPO_EDAD","SEXO","POP_STD")],
            by = c("GRUPO_EDAD","SEXO")) %>%
  mutate(
    rates = n / total_pop,
    DSR_pop = POP_STD * rates
  )

# Se suman las DSR por (año, sexo)
HOSP_C15_C26_DSR <- HOSP_C15_C26_RATES %>%
  group_by(ANO_EGRESO, SEXO) %>%
  summarise(`DSR C15-C26 Hosp. per 100,000` = sum(DSR_pop),
            .groups = 'drop')

# Tabla resumen
TABLE1 <- HOSP %>%
  group_by(ANO_EGRESO, SEXO) %>%
  summarise(N = n(), .groups = 'drop') %>%
  bind_rows(
    summarise(.,
              ANO_EGRESO = "Total",
              # Ojo aquí: sum de N por sexo
              N = sum(N),
              SEXO = unique(SEXO)) # Maneja con cuidado la agrupación
  ) %>%
  left_join(
    HOSP %>%
      filter(grepl("^C1[5-9]|^C2[0-6]", DIAG1)) %>%
      group_by(ANO_EGRESO, SEXO) %>%
      summarise(`C15-C26 Hosp.` = n(), .groups = 'drop') %>%
      bind_rows(
        summarise(.,
                  ANO_EGRESO = "Total",
                  `C15-C26 Hosp.` = sum(`C15-C26 Hosp.`),
                  SEXO = unique(SEXO))
      ),
    by = c("ANO_EGRESO","SEXO")
  ) %>%
  left_join(HOSP_C15_C26_DSR, by = c("ANO_EGRESO","SEXO")) %>%
  rename(
    Year = ANO_EGRESO,
    `N of Hospitalization` = N
  )

################################################################################
# NATIONAL
################################################################################
HOSP_C15_C26_NAT <- HOSP %>%
  filter(grepl("^C1[5-9]|^C2[0-6]", DIAG1)) %>%
  mutate(DIAG1 = substr(DIAG1, 1, 3)) %>%
  group_by(ANO_EGRESO, GRUPO_EDAD, SEXO, DIAG1) %>%
  summarise(n = n(), .groups = 'drop') %>%
  drop_na(DIAG1) %>%
  complete(
    ANO_EGRESO,
    GRUPO_EDAD,
    SEXO,
    DIAG1 = c("C15","C16","C17","C18","C19","C20",
              "C21","C22","C23","C24","C25","C26"),
    fill = list(n = 0)
  ) %>%
  mutate(across(everything(), ~na_if(., NA) %>% replace_na(0))) %>%
  mutate(ANO_EGRESO = as.character(ANO_EGRESO))

# Totales nacionales por edad, sexo y año
TOTALS_HOSP_C15_C26_NAT <- HOSP %>%
  group_by(GRUPO_EDAD, SEXO, ANO_EGRESO) %>%
  summarise(total_pop = n(), .groups = 'drop')

HOSP_C15_C26_RATES_NAT <- HOSP_C15_C26_NAT %>%
  left_join(TOTALS_HOSP_C15_C26_NAT, 
            by = c("GRUPO_EDAD","SEXO","ANO_EGRESO")) %>%
  left_join(STDPOPULATION[, c("GRUPO_EDAD","SEXO","POP_STD")],
            by = c("GRUPO_EDAD","SEXO")) %>%
  mutate(
    rates = n / total_pop,
    DSR_pop = POP_STD * rates
  )

HOSP_C15_C26_DSR_NAT <- HOSP_C15_C26_RATES_NAT %>%
  group_by(ANO_EGRESO, SEXO, DIAG1) %>%
  summarise(DSR_HOSP = sum(DSR_pop),
            .groups = 'drop')

################################################################################
# REGIONAL
################################################################################
HOSP_C15_C26_REG <- HOSP %>%
  filter(grepl("^C1[5-9]|^C2[0-6]", DIAG1)) %>%
  mutate(DIAG1 = substr(DIAG1, 1, 3)) %>%
  group_by(ANO_EGRESO, GLOSA_REGION_RESIDENCIA, GRUPO_EDAD, SEXO, DIAG1) %>%
  summarise(n = n(), .groups = 'drop') %>%
  drop_na(DIAG1) %>%
  complete(
    ANO_EGRESO,
    GLOSA_REGION_RESIDENCIA,
    GRUPO_EDAD,
    SEXO,
    DIAG1 = c("C15","C16","C17","C18","C19","C20",
              "C21","C22","C23","C24","C25","C26"),
    fill = list(n = 0)
  ) %>%
  mutate(across(everything(), ~na_if(., NA) %>% replace_na(0))) %>%
  mutate(ANO_EGRESO = as.character(ANO_EGRESO))

# Totales regionales: se completan filas "Total"
HOSP_C15_C26_REG_RATES_TOTAL <- HOSP %>%
  filter(grepl("^C1[5-9]|^C2[0-6]", DIAG1)) %>%
  mutate(DIAG1 = substr(DIAG1, 1, 3)) %>%
  group_by(GLOSA_REGION_RESIDENCIA, GRUPO_EDAD, SEXO, DIAG1) %>%
  summarise(n = n(), .groups = 'drop') %>%
  drop_na(DIAG1) %>%
  complete(
    GLOSA_REGION_RESIDENCIA,
    GRUPO_EDAD,
    SEXO,
    DIAG1 = c("C15","C16","C17","C18","C19","C20",
              "C21","C22","C23","C24","C25","C26"),
    fill = list(n = 0)
  ) %>%
  mutate(across(everything(), ~na_if(., NA) %>% replace_na(0))) %>%
  mutate(ANO_EGRESO = "Total")

# Se unen para tener año + "Total"
HOSP_C15_C26_REG <- bind_rows(HOSP_C15_C26_REG, HOSP_C15_C26_REG_RATES_TOTAL)

# Totales regionales de población
TOTALS_HOSP_C15_C26_REG <- HOSP %>%
  group_by(GRUPO_EDAD, SEXO, GLOSA_REGION_RESIDENCIA, ANO_EGRESO) %>%
  summarise(total_pop = n(), .groups = 'drop')

TOTALS_HOSP_C15_C26_REG_TOTAL <- HOSP %>%
  group_by(GRUPO_EDAD, SEXO, GLOSA_REGION_RESIDENCIA) %>%
  summarise(total_pop = n(), .groups = 'drop') %>%
  mutate(ANO_EGRESO = "Total")

# Se unen para año normal + "Total"
TOTALS_HOSP_C15_C26_REG <- bind_rows(TOTALS_HOSP_C15_C26_REG, 
                                     TOTALS_HOSP_C15_C26_REG_TOTAL)

# Cálculo de tasas regionales por edad y sexo
HOSP_C15_C26_RATES_REG <- HOSP_C15_C26_REG %>%
  left_join(TOTALS_HOSP_C15_C26_REG, 
            by = c("GRUPO_EDAD","SEXO","ANO_EGRESO","GLOSA_REGION_RESIDENCIA")) %>%
  left_join(STDPOPULATION[, c("GRUPO_EDAD","SEXO","POP_STD")],
            by = c("GRUPO_EDAD","SEXO")) %>%
  mutate(
    rates = n / total_pop,
    DSR_pop = POP_STD * rates
  ) %>%
  filter(!GLOSA_REGION_RESIDENCIA %in% c("Ignorada", "Extranjero"))

# Se agrupa por año, región y sexo para sumar el total de DSR
HOSP_C15_C26_DSR_REG <- HOSP_C15_C26_RATES_REG %>%
  group_by(ANO_EGRESO, GLOSA_REGION_RESIDENCIA, SEXO, DIAG1) %>%
  summarise(DSR_HOSP = sum(DSR_pop), .groups = 'drop') %>%
  filter(!GLOSA_REGION_RESIDENCIA %in% c("Ignorada", "Extranjero"))

################################################################################
# REGIONAL TOTAL (equivalente a "REGIONAL TOTAL C5"), pero con sexo
################################################################################
HOSP_C15C26_REG_TOT <- HOSP %>%
  filter(grepl("^C1[5-9]|^C2[0-6]", DIAG1)) %>%
  mutate(DIAG1 = substr(DIAG1, 1, 3)) %>%
  group_by(ANO_EGRESO, GLOSA_REGION_RESIDENCIA, GRUPO_EDAD, SEXO) %>%
  summarise(n = n(), .groups = 'drop') %>%
  complete(
    ANO_EGRESO,
    GLOSA_REGION_RESIDENCIA,
    GRUPO_EDAD,
    SEXO,
    fill = list(n = 0)
  ) %>%
  mutate(across(everything(), ~na_if(., NA) %>% replace_na(0))) %>%
  mutate(ANO_EGRESO = as.character(ANO_EGRESO))

HOSP_C15C26_REG_TOT_RATES_TOTAL <- HOSP %>%
  filter(grepl("^C1[5-9]|^C2[0-6]", DIAG1)) %>%
  mutate(DIAG1 = substr(DIAG1, 1, 3)) %>%
  group_by(GLOSA_REGION_RESIDENCIA, GRUPO_EDAD, SEXO) %>%
  summarise(n = n(), .groups = 'drop') %>%
  complete(
    GLOSA_REGION_RESIDENCIA,
    GRUPO_EDAD,
    SEXO,
    fill = list(n = 0)
  ) %>%
  mutate(across(everything(), ~na_if(., NA) %>% replace_na(0))) %>%
  mutate(ANO_EGRESO = "Total")

HOSP_C15C26_REG_TOT <- bind_rows(HOSP_C15C26_REG_TOT, 
                                 HOSP_C15C26_REG_TOT_RATES_TOTAL)

TOTALS_HOSP_C15C26_REG_TOT <- HOSP %>%
  group_by(GRUPO_EDAD, SEXO, GLOSA_REGION_RESIDENCIA, ANO_EGRESO) %>%
  summarise(total_pop = n(), .groups = 'drop')

TOTALS_HOSP_C15C26_REG_TOT_TOTAL <- HOSP %>%
  group_by(GRUPO_EDAD, SEXO, GLOSA_REGION_RESIDENCIA) %>%
  summarise(total_pop = n(), .groups = 'drop') %>%
  mutate(ANO_EGRESO = "Total")

TOTALS_HOSP_C15C26_REG_TOT <- bind_rows(TOTALS_HOSP_C15C26_REG_TOT, 
                                        TOTALS_HOSP_C15C26_REG_TOT_TOTAL)

# Tasas totales regionales (para todos los C15–C26) por edad y sexo
HOSP_C15C26_RATES_REG_TOT <- HOSP_C15C26_REG_TOT %>%
  left_join(TOTALS_HOSP_C15C26_REG_TOT, 
            by = c("GRUPO_EDAD","SEXO","ANO_EGRESO","GLOSA_REGION_RESIDENCIA")) %>%
  left_join(STDPOPULATION[, c("GRUPO_EDAD","SEXO","POP_STD")],
            by = c("GRUPO_EDAD","SEXO")) %>%
  mutate(
    rates = n / total_pop,
    DSR_pop = POP_STD * rates
  ) %>%
  filter(!GLOSA_REGION_RESIDENCIA %in% c("Ignorada", "Extranjero"))

HOSP_C15C26_DSR_REG_TOT <- HOSP_C15C26_RATES_REG_TOT %>%
  group_by(ANO_EGRESO, GLOSA_REGION_RESIDENCIA, SEXO) %>%
  summarise(DSR_HOSP = sum(DSR_pop), .groups = 'drop') %>%
  filter(!GLOSA_REGION_RESIDENCIA %in% c("Ignorada", "Extranjero"))

################################################################################
# Recode the DIAG1 column usando icd10_mapping también con la dimensión sexo
################################################################################
# Para el data frame con tasas nacionales por DIAG1
HOSP_C15_C26_RATES_NAT <- HOSP_C15_C26_RATES_NAT %>%
  mutate(DIAG1_full = recode(DIAG1, !!!icd10_mapping))

HOSP_C15_C26_DSR_NAT <- HOSP_C15_C26_DSR_NAT %>%
  mutate(DIAG1_full = recode(DIAG1, !!!icd10_mapping))

# Para el data frame con tasas regionales por DIAG1
HOSP_C15_C26_RATES_REG <- HOSP_C15_C26_RATES_REG %>%
  mutate(DIAG1_full = recode(DIAG1, !!!icd10_mapping))

HOSP_C15_C26_DSR_REG <- HOSP_C15_C26_DSR_REG %>%
  mutate(DIAG1_full = recode(DIAG1, !!!icd10_mapping))
```

# Results

Figure 1 and Table 1 presents the annual trends in standardized hospitalization rates attributed to DMNC (ICD codes C15-C26) per 100,000 hospitalized individuals analyzed from 2010 to 2023. The data were stratified by sex to allow comparison of patterns between women and men.

In the panel corresponding to females, colon and stomach neoplasms displayed relatively high hospitalization rates throughout the period. A pronounced peak was observed in the hospitalization rate for rectal cancer in 2016, surpassing 180 hospitalizations per 100,000. The rates for colon cancer show an increasing trend, from 117 per 100,000 hospitalized patients in 2010 to 180 per 100,000 hospitalized patients. Stomach cancer is also a significant cause of hospitalization, although it has lower rates than colon cancer. Other cancers, such as those of the pancreas, liver, and intrahepatic bile ducts, show intermediate rates, while cancers of the esophagus, small intestine, and anus register considerably lower rates.

For males, hospitalization rates due to colon cancer are consistently the highest among the specific types charted, showing an overall upward trend from 2010, with a peak close to 181 per 100,000 in 2023, far exceeding the rates observed in women for this type of cancer. Stomach cancer also displays high rates in men, generally higher than those observed in women, with a relatively stable trend of 170-180 per 100,000 hospitalized patients between 2010 and 2020, dropping to 148 per 100,000 hospitalized patients in 2023. Liver and intrahepatic bile duct cancers, as well as pancreatic and rectal cancers, are also relevant causes of hospitalization in men, with the latter showing a notable peak near 145 per 100,000 in 2016. As with women, cancers of the esophagus, small intestine, and anus had the lowest rates.

```{r}

datatable(TABLE1 %>%
  mutate(across(everything(), ~ ifelse(is.na(.), "", .))) %>% 
    mutate(Year = as.factor(Year),
           SEXO = as.factor(SEXO)),
  filter = 'top', 
  options = list(
    pageLength = 20,
    scrollX = TRUE,
    scrollY = "800px",
    autoWidth = TRUE)# Agrupa usando las dos primeras columnas
  )      # Continue adding for each column you want to make bold
```

```{r}
#| fig-width: 10
#| fig-height: 8 
#| fig-cap: "Figure 1: Standardized hospitalization rates for malignant neoplasms of digestive organs (ICD C15-C26) per 100,000 hospitalized patients stratified by sex (Female and Male), during the period 2010-2023. The lines represent different types of digestive cancer (see the legend)."
#
ggplot(HOSP_C15_C26_DSR_NAT, 
       aes(x = as.factor(ANO_EGRESO), 
           y = DSR_HOSP, 
           color = DIAG1_full, 
           group = DIAG1_full)) +
  geom_line(size = 0.8, alpha = 0.5) +
  geom_text(aes(label = round(DSR_HOSP, 1)), 
            vjust = -0.5, 
            size = 3.5, 
            show.legend = FALSE) +
  labs(
    x = "Year",
    y = "Standardized rate of malignant neoplasm hospitalizations (C15–C26) per 100,000 hospitalized",
    color = "Malignant neoplasm (Digestive organs)"
  ) +
  theme_minimal() +
  # Colores de ejemplo para cada neoplasia de C15–C26
  scale_color_manual(values = c(
    "Esophagus"                                   = "red",
    "Stomach"                                     = "orange",
    "Small intestine"                             = "green",
    "Colon"                                       = "blue",
    "Rectosigmoid junction"                       = "purple",
    "Rectum"                                      = "cyan",
    "Anus and anal canal"                         = "pink",
    "Liver and intrahepatic bile ducts"           = "brown",
    "Gallbladder"                                 = "magenta",
    "Other parts of biliary tract"                = "gray",
    "Pancreas"                                    = "darkgreen",
    "Other and ill-defined digestive organs"      = "darkblue"
  )) +
  facet_wrap(~SEXO) +  # Facetas por sexo
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  )

# Guardar la figura
ggsave("fig1.png", dpi = 1200, width = 10, height = 8)

```
Figure 2 shows the distribution of standardized hospitalization rates for DMNC (ICD C15-C26) according to age and sex. The rate was expressed per 100,000 hospitalized individuals. The data are presented using points indicating the specific rate for each type of digestive cancer within ten-year age intervals (0-9, 10-19, ..., 90+ years), separating the results for women and men.

The most prominent pattern observed in both sexes was a substantial and progressive increase in hospitalization rates with age for the vast majority of digestive cancers analyzed. The rates were minimal or close to zero in the youngest age groups (0-39 years). Starting from the to 40-49 age group, a notable rise began, which became more pronounced in the subsequent age groups.

For females, it was observed that the highest rates in the older age groups (between 50 and 79 years) corresponded to colon cancer, followed by stomach cancer. Pancreatic, rectal, and liver/intrahepatic bile duct cancers also show considerable increases with advancing age.

In males, colon cancer shows the highest rates in most adult and elderly age groups, reaching values above 500 per 100,000 in the 50-59 and 60-69 age groups. These rates are markedly higher than those observed in women in the same age groups. Stomach cancer also presents high rates, surpassing those seen in females, especially in older ages.

In both sexes, neoplasms such as those of the esophagus, small intestine, gallbladder, rectum, and rectosigmoid junction present significantly lower hospitalization rates compared to those previously mentioned, although they also tend to increase in the older age groups.

```{r}
#| fig-height: 8
#| fig-width: 16
#| fig-cap: "Figure 2: Age-standardized hospitalization rates for malignant neoplasms of the digestive organs (ICD C15-C26) per 100,000 hospitalized patients, stratified by age group and sex (Female and Male), accumulated between 2010 and 2023. Each point represents the rate of a specific type of cancer (as indicated in the legend) in a given age group."
#| 
HOSP_C15_C26_RATES_NAT %>%
  # 1) Agrupación por edad, diagnóstico y sexo
  group_by(GRUPO_EDAD, DIAG1_full, SEXO) %>%
  summarise(DSR_total = sum(DSR_pop), .groups = "drop") %>%
  
  # 2) Gráfico de puntos con facetas por sexo
  ggplot(aes(
    x = GRUPO_EDAD,
    y = DSR_total,
    shape = DIAG1_full,
    color = DIAG1_full
  )) +
  geom_point(size = 4, alpha = 0.6) +
  
  # 3) Etiquetas de valores en cada punto
  geom_text(
    aes(label = round(DSR_total, 1)),
    vjust = -0.5,
    size = 3.5,
    show.legend = FALSE
  ) +
  
  # 4) Escalas manuales de forma y color
  scale_shape_manual(values = c(
    "Esophagus"                                   = 16,
    "Stomach"                                     = 17,
    "Small intestine"                             = 15,
    "Colon"                                       = 3,
    "Rectosigmoid junction"                       = 5,
    "Rectum"                                      = 8,
    "Anus and anal canal"                         = 18,
    "Liver and intrahepatic bile ducts"           = 7,
    "Gallbladder"                                 = 13,
    "Other parts of biliary tract"                = 11,
    "Pancreas"                                    = 9,
    "Other and ill-defined digestive organs"      = 10
  )) +
  scale_color_manual(values = c(
    "Esophagus"                                   = "red",
    "Stomach"                                     = "orange",
    "Small intestine"                             = "green",
    "Colon"                                       = "blue",
    "Rectosigmoid junction"                       = "purple",
    "Rectum"                                      = "cyan",
    "Anus and anal canal"                         = "pink",
    "Liver and intrahepatic bile ducts"           = "brown",
    "Gallbladder"                                 = "magenta",
    "Other parts of biliary tract"                = "darkgray",
    "Pancreas"                                    = "darkgreen",
    "Other and ill-defined digestive organs"      = "darkblue"
  )) +
  
  # 5) Etiquetas y tema
  labs(
    x = "Age Group",
    y = "Standardized rate of malignant neoplasm hospitalizations (C15–C26) per 100,000 hospitalized",
    color = "Malignant neoplasm (Digestive organs)",
    shape = "Malignant neoplasm (Digestive organs)"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  ) +
  
  # 6) Para mantener el orden de GRUPO_EDAD y ejes
  scale_x_discrete(limits = levels(HOSP_C15_C26_RATES_NAT$GRUPO_EDAD)) +
  scale_y_continuous(breaks = seq(0, 1500, by = 100)) +
  coord_flip() +
  
  # 7) Facetado por sexo
  facet_wrap(~SEXO)

# Guardar la figura
ggsave("fig2.png", dpi = 1200, width = 10, height = 8)
```
Figure 3 shows the adjusted hospitalization rates for specific digestive cancers (esophagus, stomach, colorectal, liver, and pancreas) broken down by sex and administrative region of Chile. There is marked variability in the rates, with notable geographic differences and a clear predominance of higher rates in men than in women in all regions.

There was considerable heterogeneity in hospitalization rates between regions for most of the digestive cancers studied. Consistently, across most regions, hospitalization rates for colon and stomach cancers are higher in men than in women. However, no uniform pattern exists at the national level. Instead, certain regions display distinctly different hospitalization burdens for specific types of cancer.

Among women, the rates are considerably lower than those among men, although hospitalization for colorectal and gastric cancer remains significant. The regions of Arica and Parinacota (316 per 100,000 hospitalizations), Valparaíso (192.5 cases), Los Lagos (186.1), Metropolitana (162.6 cases), followed by Araucanía (141.9 cases), and Los Ríos (119.4 cases) present the highest rates of colon cancer among women. The consistently lower rates in women than in men were observed across all regions and for all evaluated types of digestive cancer.

Among men, hospitalization rates are particularly high for gastric, colorectal, and rectal cancer. The regions of Arica and Parinacota (307 cases per 100,000 hospitalizations), Los Lagos (319.4), Araucanía (230.6 cases), and Biobío (201.6 cases) had the highest rates of colon or stomach cancer, indicating a significant hospital burden from these cancers. Similarly, the Metropolitana and Valparaíso regions showed elevated rates (165.8 and 180.1 cases respectively).



```{r}
#| fig-height: 10
#| fig-width: 16
#| fig-cap: "Figure 3. Standardized hospitalization rates for malignant neoplasms of digestive organs (ICD C15-C26) per 100,000 hospitalized individuals, stratified by the geographic region of Chile and sex (female and male). Each point represents the rate of a specific cancer type (as indicated in the legend) within a given region."

# Define the order of the regions from north to south
region_order <- c(
  "De Arica y Parinacota", 
  "De Tarapacá", 
  "De Antofagasta", 
  "De Atacama", 
  "De Coquimbo", 
  "De Valparaíso", 
  "Metropolitana de Santiago", 
  "Del Libertador B. O'Higgins", 
  "Del Maule", 
  "Del Bíobío", 
  "De Ñuble", 
  "De La Araucanía", 
  "De Los Ríos", 
  "De Los Lagos", 
  "De Aisén del Gral. C. Ibáñez del Campo", 
  "De Magallanes y de La Antártica Chilena"
)

# Apply the order to the GLOSA_REGION_RESIDENCIA column
HOSP_C15_C26_DSR_REG <- HOSP_C15_C26_DSR_REG %>%
  mutate(GLOSA_REGION_RESIDENCIA = factor(GLOSA_REGION_RESIDENCIA, levels = region_order))
# Plot
HOSP_C15_C26_DSR_REG %>%
  filter(ANO_EGRESO == "Total") %>%
  ggplot(aes(x = GLOSA_REGION_RESIDENCIA, 
             y = DSR_HOSP, 
             shape = DIAG1_full, 
             color = DIAG1_full)) +
  # Puntos
  geom_point(size = 4, alpha = 0.6) +
  # Etiquetas de la tasa en cada shape
  geom_text(aes(label = round(DSR_HOSP, 1)),
            vjust = -0.5,
            size = 3.5,
            show.legend = FALSE) +
  # Escala de formas para cada neoplasia digestiva
  scale_shape_manual(values = c(
    "Esophagus"                                   = 16,  # Circle
    "Stomach"                                     = 17,  # Triangle up
    "Small intestine"                             = 15,  # Square
    "Colon"                                       = 3,   # Plus
    "Rectosigmoid junction"                       = 5,   # Cross
    "Rectum"                                      = 8,   # Star
    "Anus and anal canal"                         = 18,  # Diamond
    "Liver and intrahepatic bile ducts"           = 7,   # Triangle down
    "Gallbladder"                                 = 13,  
    "Other parts of biliary tract"                = 11,
    "Pancreas"                                    = 9,
    "Other and ill-defined digestive organs"      = 10
  )) +
  # Escala de colores para cada neoplasia digestiva
  scale_color_manual(values = c(
    "Esophagus"                                   = "red",
    "Stomach"                                     = "orange",
    "Small intestine"                             = "green",
    "Colon"                                       = "blue",
    "Rectosigmoid junction"                       = "purple",
    "Rectum"                                      = "cyan",
    "Anus and anal canal"                         = "pink",
    "Liver and intrahepatic bile ducts"           = "brown",
    "Gallbladder"                                 = "magenta",
    "Other parts of biliary tract"                = "darkgray",
    "Pancreas"                                    = "darkgreen",
    "Other and ill-defined digestive organs"      = "darkblue"
  )) +
  # Facetas por Sexo
  facet_wrap(~SEXO) +
  labs(
    x = "Region", 
    y = "Standardized rate of malignant neoplasm hospitalizations (C15–C26) per 100,000 hospitalized",
    color = "Malignant neoplasm (Digestive organs)",
    shape = "Malignant neoplasm (Digestive organs)"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "top"
  ) +
  # Orden inverso en el eje X (si quieres de sur a norte o viceversa)
  scale_x_discrete(limits = rev(region_order)) +
  # Ajusta los límites y saltos de acuerdo a tus datos
  scale_y_continuous(breaks = seq(0, 500, by = 50)) +
  # Voltea los ejes para listar las regiones verticalmente
  coord_flip()

ggsave("fig3.png", dpi = 1200, width = 12, height = 8)
```

```{r}
#| fig-height: 10
#| fig-width: 10
#| include: false

# Load the regional map of Chile from chilemapas
chile_regions <- generar_regiones(mapa = chilemapas::mapa_comunas) %>%
  mutate(
    region_name = recode(codigo_region,
      "01" = "De Tarapacá",
      "02" = "De Antofagasta",
      "03" = "De Atacama",
      "04" = "De Coquimbo",
      "05" = "De Valparaíso",
      "06" = "Del Libertador B. O'Higgins",
      "07" = "Del Maule",
      "08" = "Del Bíobío",
      "09" = "De La Araucanía",
      "10" = "De Los Lagos",
      "11" = "De Aisén del Gral. C. Ibáñez del Campo",
      "12" = "De Magallanes y de La Antártica Chilena",
      "13" = "Metropolitana de Santiago",
      "14" = "De Los Ríos",
      "15" = "De Arica y Parinacota",
      "16" = "De Ñuble"
    )
  )

# Resumir (sumar) los valores de DSR_HOSP para cada región
HOSP_C15_C26_DSR_REG_sum <- HOSP_C15_C26_DSR_REG %>%
  filter(ANO_EGRESO == "Total", SEXO == "Female") %>%
  group_by(GLOSA_REGION_RESIDENCIA) %>%
  summarise(DSR_HOSP = sum(DSR_HOSP, na.rm = TRUE))

# Unir con el mapa de Chile y preparar la geometría
plot_data_sum <- HOSP_C15_C26_DSR_REG_sum %>%
  left_join(chile_regions, by = c("GLOSA_REGION_RESIDENCIA" = "region_name")) %>%
  mutate(centroid = st_centroid(geometry))

# Graficar en un solo mapa (sin facet_wrap)
ggplot(plot_data_sum) + 
  geom_sf(aes(fill = DSR_HOSP, geometry = geometry), 
          color = "white", size = 0.3) +
  geom_segment(aes(x = st_coordinates(centroid)[, 1],
                   y = st_coordinates(centroid)[, 2],
                   xend = st_coordinates(centroid)[, 1] - 2,
                   yend = st_coordinates(centroid)[, 2]),
               color = "turquoise4", size = 0.5, 
               arrow = arrow(length = unit(0.1, "inches"))) +
  geom_text(aes(label = GLOSA_REGION_RESIDENCIA,
                x = st_coordinates(centroid)[, 1] - 2,
                y = st_coordinates(centroid)[, 2]),
            size = 3, color = "purple4", hjust = 1) +
  scale_fill_viridis_c(option = "D", alpha = 0.8,
                       name = "Standardized rate of malignant neoplasm\nhospitalizations per 100,000 hospitalized females.") +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  ) +
  coord_sf(xlim = c(-90, -65), ylim = c(-55, -15))

ggsave("fig4.png", dpi = 1200, width = 10, height = 16)
```
```{r}
#| fig-height: 10
#| fig-width: 10
#| include: false
# Resumir (sumar) los valores de DSR_HOSP para cada región
HOSP_C15_C26_DSR_REG_sum <- HOSP_C15_C26_DSR_REG %>%
  filter(ANO_EGRESO == "Total", SEXO == "Male") %>%
  group_by(GLOSA_REGION_RESIDENCIA) %>%
  summarise(DSR_HOSP = sum(DSR_HOSP, na.rm = TRUE))

# Unir con el mapa de Chile y preparar la geometría
plot_data_sum <- HOSP_C15_C26_DSR_REG_sum %>%
  left_join(chile_regions, by = c("GLOSA_REGION_RESIDENCIA" = "region_name")) %>%
  mutate(centroid = st_centroid(geometry))

# Graficar en un solo mapa (sin facet_wrap)
ggplot(plot_data_sum) + 
  geom_sf(aes(fill = DSR_HOSP, geometry = geometry), 
          color = "white", size = 0.3) +
  geom_segment(aes(x = st_coordinates(centroid)[, 1],
                   y = st_coordinates(centroid)[, 2],
                   xend = st_coordinates(centroid)[, 1] - 2,
                   yend = st_coordinates(centroid)[, 2]),
               color = "turquoise4", size = 0.5, 
               arrow = arrow(length = unit(0.1, "inches"))) +
  geom_text(aes(label = GLOSA_REGION_RESIDENCIA,
                x = st_coordinates(centroid)[, 1] - 2,
                y = st_coordinates(centroid)[, 2]),
            size = 3, color = "purple4", hjust = 1) +
  scale_fill_viridis_c(option = "D", alpha = 0.8,
                       name = "Standardized rate of malignant neoplasm\nhospitalizations per 100,000 hospitalized males.") +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5),
    legend.position = "right"
  ) +
  coord_sf(xlim = c(-90, -65), ylim = c(-55, -15))

ggsave("fig5.png", dpi = 1200, width = 10, height = 16)
```

```{r}
#| fig-height: 10
#| fig-width: 14
# Combinar datos para ambos sexos
HOSP_C15_C26_DSR_REG_combined <- HOSP_C15_C26_DSR_REG %>%
  filter(ANO_EGRESO == "Total") %>%
  group_by(GLOSA_REGION_RESIDENCIA, SEXO) %>%
  summarise(DSR_HOSP = sum(DSR_HOSP, na.rm = TRUE))

# Unir con el mapa y preparar geometrías
plot_data_both <- HOSP_C15_C26_DSR_REG_combined %>%
  left_join(chile_regions, by = c("GLOSA_REGION_RESIDENCIA" = "region_name")) %>%
  mutate(centroid = st_centroid(geometry))

# Crear el gráfico con facetas - etiquetas solo en Female
ggplot(plot_data_both) + 
  geom_sf(aes(fill = DSR_HOSP, geometry = geometry), 
          color = "white", size = 0.3) +
  # Mostrar segmentos solo para Female
  geom_segment(data = filter(plot_data_both, SEXO == "Female"),
               aes(x = st_coordinates(centroid)[, 1],
                   y = st_coordinates(centroid)[, 2],
                   xend = st_coordinates(centroid)[, 1] - 2,
                   yend = st_coordinates(centroid)[, 2]),
               color = "turquoise4", size = 0.5, 
               arrow = arrow(length = unit(0.1, "inches"))) +
  # Mostrar texto solo para Female
  geom_text(data = filter(plot_data_both, SEXO == "Female"),
            aes(label = GLOSA_REGION_RESIDENCIA,
                x = st_coordinates(centroid)[, 1] - 2,
                y = st_coordinates(centroid)[, 2]),
            size = 2.5, color = "purple4", hjust = 1) +
  scale_fill_viridis_c(option = "D", alpha = 0.8,
                       name = "Standardized rate per 100,000") +
  labs(title = "Standardized rate of malignant neoplasm hospitalizations by region and sex",
       x = NULL, y = NULL) +
  facet_wrap(~ SEXO, ncol = 2) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank(),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 16),
    legend.position = "bottom",
    legend.key.width = unit(2, "cm"),
    strip.text = element_text(size = 14, face = "bold"),
    panel.spacing = unit(0, "lines")
  ) +
  coord_sf(xlim = c(-90, -65), ylim = c(-55, -15))

ggsave("fig_combined_facets.png", dpi = 1200, width = 14, height = 10)
```

# GitHub Repository

This code is automatically replicated with the data. The complete code for this analysis, including the statistical models, visualizations, and diagnostics, is available in the following [GitHub repository](https://github.com/AmaruSimonAgueroJimenez/Malignant-neoplasms-of-digestive-organs-C15-C26-Chile-2010-2023). You can also access it using the following QR code.

```{r}
#| fig-height: 4
#| fig-width: 4
#| fig-align: center
#| results: 'asis'
# URL del repositorio
repo_url <- "https://github.com/AmaruSimonAgueroJimenez/Malignant-neoplasms-of-digestive-organs-C15-C26-Chile-2010-2023"

# Generar código QR
qr <- qr_code(repo_url)
plot(qr)

# png("qr1.png", width = 6000, height = 6000, res = 300)
# plot(qr)
# dev.off()
```


The report can be found at [this address](https://amarusimonaguerojimenez.github.io/Malignant-neoplasms-of-digestive-organs-C15-C26-Chile-2010-2023/). You can also access it using the following QR code.

```{r}
#| fig-height: 4
#| fig-width: 4
#| fig-align: center
#| results: 'asis'
# URL del repositorio
repo_url <- "https://amarusimonaguerojimenez.github.io/Malignant-neoplasms-of-digestive-organs-C15-C26-Chile-2010-2023/"

# Generar código QR
qr <- qr_code(repo_url)
plot(qr)

# png("qr2.png", width = 6000, height = 6000, res = 300)
# plot(qr)
# dev.off()
```


